Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-11 paper/sections/implementation.tex
[ paper ] implementation added, some proofs added.

\section{Proof-Reconstruction}
\label{sec:proof-reconstruction}

\subsection{Workflow}
\label{ssec:workflow}

\begin{figure}
\centering

\begin{tikzpicture}[scale=0.9]
\node[text width=2cm, align=center](problem) at (0,0)
  {1. CPL \\ Problem.};

\node[right = 1.1cm of problem, text width=2cm, align=center]
  (tptp){2. \TPTP \\ Problem.};

\node[right= 1.1cm of tptp, text width=2cm, align=center]
   (metis) {3. \Metis \\ Prover.};

\node[right= 1.1cm of metis, text width=2cm, align=center, inner sep=10pt]
   (tstp) {4. \TSTP \\ Derivation.};

\node[below= 0.5cm of tstp, text width=2cm, align=center, inner sep=10pt]
  (athena) {5. \Haskell \\ Traslator.};

\node[left = 1.1cm of athena, text width=2cm, align=center]
   (agdafile) {6. \Agda \\ Proof.};

\node[left = 1.1cm of agdafile, text width=2cm, align=center]
   (agda) {7. Proof Checking.};

\node[below = 0.5cm of problem, text width=2cm, align=center]
   (verified) {8.1 Proof \\ Checked.};

\node[below = 0.5cm of verified, text width=2cm, align=center]
   (failure) {8.2 Failure.};

% node[below] {send to}
\draw[->, thick] (problem) to
  % node[below] {\tiny encoding}
  (tptp);
\draw[->, thick] (tptp) to
  % node[below] {\tiny }
  (metis);
\draw[->, thick] (metis) to
  % node[below] {\tiny replies on}
  (tstp);
\draw[->, thick] (tstp) to
  % node[right] {\tiny parsing}
  (athena);
\draw[->, thick] (athena) to
  % node[below] {\tiny traslation}
  (agdafile);
\draw[->, thick] (agdafile) to
  % node[below] {\tiny type-checking}
  (agda);
\draw[->, thick] (agda) to (verified);
\draw[->, thick, gray] (agda) to (failure);
\end{tikzpicture}
\caption{proof-reconstruction overview.}
\label{fig:proof-reconstruction-workflow}
\end{figure}

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by Sultana in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is the proof
assistant, \Agda. We choose \Agda, but another proof assistant with
support of type theory and features similar like \Agda is valid; our
results do not depend on the proof assistant but the support of
inductive data types, termination checking, and type-checking.

Following the diagram in Fig.
\ref{fig:proof-reconstruction-workflow}, the process can begin with
a problem in \CPL that we encode in \TPTP format to be sent as the
input of the \Metis prover.
The proof-search algorithms of \Metis in step No. 3 perform a
search for a solution; if they success, \Metis
replies a derivation in \TSTP format of the proof, else \Metis may
no stop running forever or replies the problem is counter feasible.

With such a derivation, we process it with our \Haskell translator
tool, \Athena~\cite{Athena}.
\Athena parses the \TSTP format, analyze the
derivation and generates a representation of the natural deduction
proof using a version of a tree data structure (see the properties
of this tree in subsection \ref{ssec:metis-proofs}). As a result,
\Athena generates an \Agda file with such a proof. We have included
in this file, imports for \Agda libraries that give the support for
the logic~\cite{AgdaProp}, and versions of the \Metis' inference
rules~\cite{AgdaMetis}.

Finally, we use the proof assistant in step No. 7 to type-check the
proof and verify its correctness. Next, we only have one answer:
type-checking succeed (i.e., the proof is valid) or type-checking failed.
If the type-check fails, one of the three actors in the process are
responsible: \Metis, \Athena, or \Agda. We are responsible for
errors committed in the translation from \TSTP to \Agda; lack of the
coverage of the \Metis' inference rules, or bad assumptions about
them. \improvement{Make this statement stronger and clear.}

In the rest of this section, we provide a formal description using
type theory to build definitions and theorems for the functions
necessary to emulate the \Metis' inference rules.

% ...................................................................

% \footnote{We use the \Agda standard library's implementation of the List data type.}.

@@

In \Agda, we define a proposition formula
as an inductive type using the keyword \texttt{data}.
The \name{PropFormula} data type stands for propositions,\cite{Altenkirch2015}.
We include every connective, logic constant as a constructor.
We also define the constructor for atomic propositions \verb!Var!
to represent based on a natural number $i$, the atomic proposition
$p_i$ for a finite universe of them ($i \leq n$), this is $Fin n$ in the
definition.

@@

\begin{remark}
In the above definition, we defined all connectives, execept negation, as
a binary connective. In \Agda, we named constructors for those functions
that inhabit the data type, for example, the boolean connecetives and
the function for atomic propositions, \verb!Var!, are constructors.
The underscores in a definition like
\begin{verbatim}
_⇔_ : (φ ψ : PropFormula) → PropFormula
\end{verbatim}
stands for an argument, and \verb!(φ ψ : PropFormula)! intend for the
uncurryfied version of the type:
$$PropFormula → PropFormula → PropFormula$$
\end{remark}

\begin{definition}(Negative Normal Form (NNF))
Una forma normal negativa para la formula $\varphi$ es aquella
en la cual negaciones aparecen solamente en los literales y
la expresión se encuentra en términos unicamente
de los operadores booleanos ($\neg$, $\wedge$, $\vee$).
\end{definition}

Para transformar una formula $\varphi$ a su forma equivalente NNF,
aplicamos recursivamente las siguientes ecuaciones que definen
la función \name{nnf}.

\begin{equation}
\label{eq:nnf-definition}
\nnf(φ) =
\begin{cases}
\nnf(φ₁) ∧ \nnf(φ₂)
   & φ := φ₁ ∧ φ₂  \\

\nnf(φ₁) ∨ \nnf(φ₂)
   & φ := φ₁ ∨ φ₂  \\

\nnf(¬ φ₁ ∨ φ₂)
   & φ := φ₁ ⇒ φ₂  \\

\nnf(φ₁ ⇒ φ₂ ∧ φ₂ ⇒ φ₁)
   & φ := φ₁ ⇔ φ₂\\

\nnf(¬ φ₁ ∨ ¬ φ₂)
   & φ := ¬ (φ₁ ∧ φ₂) \\

\nnf(¬ φ₁ ∧ ¬ φ₂)
   & φ := ¬ (φ₁ ∨ φ₂) \\

\nnf(φ₁)
   & φ := ¬ ¬ φ₁     \\

\nnf(¬ (φ₂ ∨ ¬ φ₁))
   & φ := ¬ (φ₁ ⇒ φ₂)  \\

\nnf(¬ (φ₁ ⇒ φ₂ ∧ φ₂ ⇒ φ₁))
   & φ := ¬ (φ₁ ⇔ φ₂) \\

⊥  & φ := ¬ ⊤        \\

⊤  & φ := ¬ ⊥        \\

φ  & \text{otherwise}
\end{cases}
\end{equation}

\begin{definition}
Una funcion recursiva se dice bien estructura, si el tamaño de los argumentos
con que con esta se define se reducen en tamaño por cada llamada a la
funcion.
\end{definition}

De la anterior definincion, sintacticamente la funcion nnf no es recursivamente
bien estructurado. El problem que no lo sea esta en nuestra traducion
de este tipo de funciones a la teoria de tipos.

Veamos otros ejemplos de funciones con el mismo problema que utilizaremos mas
adelante para formalizar la inferencia strip.

\begin{equation}
\label{eq:unshunt-definition}
\unshunt(φ) =
\begin{cases}
\unshunt((φ₁ ∧ φ₂) ⇒ φ₃)
  & φ := φ₁ ⇒ (φ₂ ⇒ φ₃)\\

\unshunt(φ₁ ⇒ φ₂) ∧ \unshunt(φ₁ ⇒ φ₃)
  & φ := φ₁ ⇒ (φ₂ ∧ φ₃)\\

φ &\text{otherwise}
\end{cases}
\end{equation}

Para traducir las funciones en un manera que esten bien definidas,
utilizamos el metodo propuesto por Ana Bove, el cual basicamente consiste

una funcion general recursiva $f:\sigma \to \tau$, su formalizacion en
teoria de tipos consiste en definir un predicado inductivo $Ind$,
$P:\sigma \to Ind$

En nuestro caso, el conjunto de los numeros servira como nuestro
conjunto inductivo y como segundo argumento para nueva funcion.
De manera, que la funcion g estara definida por recursion estructural en
su segundo termino.

Para aplicar la tecnica antes mencionada para nuestras funciones,
veamos su aplicación con la funcion \unshunt porque para las demas,
seguiremos un camino similar, f con tipo:

\[
\unshunt^{*} : PROP \to \mathcal{N} \to PROP
\]

\begin{equation}
\label{eq:unshunt-structured}
\unshunt^{*}(φ, n) =
\begin{cases}
\unshunt^{*}((φ₁ ∧ φ₂) ⇒ φ₃, n-1)
  & φ := φ₁ ⇒ (φ₂ ⇒ φ₃)\\

\unshunt^{*}(φ₁ ⇒ φ₂, n-1) ∧ \unshunt^{*}(φ₁ ⇒ φ₃,n-1)
  & φ := φ₁ ⇒ (φ₂ ∧ φ₃)\\

φ & n := 0
\end{cases}
\end{equation}

\begin{definition}
El rango $rank(\phi)$ de una proposition $\phi$ es definido como

\begin{equation}
\label{eq:rank-definition}
rank(\phi)=
\begin{cases}
0 &\text{for atomic }\phi\\

\max(rank(\phi_1),rank(\phi_2)) + 1
  & \phi := \phi_1\,\square\, \phi_2 \\
rank(\phi_1) + 1  & \phi := \neg \phi_1
\end{cases}
\end{equation}

\end{definition}

Then, we our final revision for the function $\unshunt$ is the
function$\unshunt^{+}$ with type $PROP \to PROP$
with a structured recursion,

\begin{equation}
\unshunt^{+}(\varphi) = \unshunt^{*}(\varphi, rank(\varphi))
\end{equation}

In the following section, our task will consist of a searching
appropriate functions like $rank$ to assure the termination of
our transformation functions like $nnf$.

% ................................................

\subsection{Splitting the goal}

\begin{equation}
\label{eq:split-definition}
\mysplit(φ) =
\begin{cases}
\unshunt(\mysplit(φ₁)) ∧ \unshunt(φ₁ ⇒ \mysplit(φ₂)
  & φ := φ₁ ∧ φ₂\\

\unshunt(¬ φ₁ ⇒ \mysplit(φ₂))
  & φ := φ₁ ∨ φ₂\\

\unshunt(φ₁ ⇒ \mysplit(φ₂))
  & φ := φ₁ ⇒ φ₂\\

\unshunt(φ₁ ⇒ \mysplit(φ₂)) ∧ \unshunt(φ₂ ⇒ \mysplit(φ₁))
  & φ := φ₁ ⇔ φ₂\\

\unshunt(φ₁ ⇒ \mysplit(¬ φ₂))
  & φ := ¬ (φ₁ ∧ φ₂)\\

\unshunt(\mysplit(¬ φ₁)) ∧ \unshunt(¬ φ₁ ⇒ \mysplit(¬ φ₂))
  & φ := ¬ (φ₁ ∨ φ₂)\\

\unshunt(\mysplit(φ₁)) ∧ \unshunt(φ₁ ⇒ \mysplit(¬ φ₂))
  & φ := ¬ (φ₁ ⇒ φ₂)\\

\unshunt(φ₁ ⇒ \mysplit(¬ φ₂)) ∧ \unshunt(¬ φ₂ ⇒ \mysplit(φ₁))
  & φ := ¬ (φ₁ ⇔ φ₂)\\

\unshunt(\mysplit(φ₁)
  & φ := ¬ ¬ φ₁\\

⊤ & φ := ¬ ⊥\\

⊥ & φ := ¬ ⊤\\

φ &\text{otherwise}
\end{cases}
\end{equation}

>>>>

\section{Implementation}
\label{sec:implementation}

@@

In \Agda, we define a proposition formula
as an inductive type using the keyword \texttt{data} extending the
its definition in \cite{Altenkirch2015}.  
The \name{PropFormula} data type stands for propositions
as we define in section \ref{sec:preliminaries}.
We include every connective and logic constant as a constructor.
We also define the constructor for atomic propositions denoted by
\verb!Var!. This atomic expressions are represented using a natural 
number $i$ bounded by a size of the universe $n$ for those atomic 
propositions (in \Agda, we use the data type \verb!Fin! from the 
standard  library for such a purpose).

@@

\begin{remark}
The underscores in the above definition denote holes for a function of arity equivalent to the number of such holes.
\end{remark}

\begin{remark}
The type notation
\begin{verbatim}
_⇔_ : (φ ψ : PropFormula) → PropFormula
\end{verbatim}
is a syntax sugar for the type
\begin{verbatim}
PropFormula → PropFormula → PropFormula.
\end{verbatim}
\end{remark}

\begin{example}
Using Unicode translating a proposition from a textbook to \Agda 
using the above data type is straightforward. 
\begin{verbatim}
r : PropFormula
p = Var (# 0)
q = Var (# 1)
r = Var (# 2)

φ : PropFormula
φ = (¬ r ∨ p) ∧ (¬ r ∨ q)
\end{verbatim}
\end{example}

### 2017-09-11 paper/sections/introduction.tex
[ paper ] implementation added, some proofs added.

proof-reconstruction is a hard labor since it depends on the
integration of two complex systems. On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories. A \SAT solver (e.g., \name{zChaff}
\cite{Moskewicz2001} and \name{MiniSat}~\cite{Een2004}) to prove
unsatisfiability of \CNF formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and \name{DepQBF}
\cite{Lonsing2017}) to prove satisfiability and invalidity of
quantified Boolean formulas, a \SMT solver (e.g \name{CVC4}
\cite{Barrett2011}, \name{veriT}~\cite{bouton2009}, and \name{Z3}
\cite{DeMoura2008}) to prove unsatisfiability of formulas from
first-order logic with theories, and a prover for validity of
formulas from first-order logic with equality (e.g., \name{E}
\cite{Schulz:AICOM-2002}, \name{leanCoP}~\cite{Otten2008}, \Metis
\cite{hurd2003first}, \name{SPASS} ~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II} \cite{Benzmuller2008} and \name{Satallax}
\cite{Brown2012}) or intuitionistic logic (e.g., \name{ileanCoP}
\cite{Otten2008}, \name{JProver} \cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam}, \name{Coq}
\cite{coqteam}, \name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems like
operators, logic variables, axioms, and theorems.

@@

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows. In section
\ref{sec:metis-language-and-proofs}, we briefly introduce the \Metis
prover. In section \ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
In section~\ref{sec:conclusions}, we discuss
some limitations, advantages, and disadvantages with other similar
proof-reconstruction tools for ending up with some conclusions and
the future work. Finally, we list some related works in proof
reconstruction in section \ref{sec:related-work}.

>>>>

Proof-reconstruction is a hard labor since it depends on the
integration of two complex systems. On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories. A \SAT solver (e.g., \name{zChaff}
\cite{Moskewicz2001} and \name{MiniSat}~\cite{Een2004}) to prove
unsatisfiability of \CNF formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and \name{DepQBF}
\cite{Lonsing2017}) to prove satisfiability and invalidity of
quantified Boolean formulas, a \SMT solver (e.g \name{CVC4}
\cite{Barrett2011}, \name{veriT}~\cite{bouton2009}, and \name{Z3}
\cite{DeMoura2008}) to prove unsatisfiability of formulas from
first-order logic with theories, and a prover for validity of
formulas from first-order logic with equality (e.g., \name{E}
\cite{Schulz:AICOM-2002}, \name{leanCoP}~\cite{Otten2008}, \Metis
\cite{hurd2003first}, \name{SPASS} ~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II} \cite{Benzmuller2008} and \name{Satallax}
\cite{Brown2012}) or intuitionistic logic (e.g., \name{ileanCoP}
\cite{Otten2008}, \name{JProver} \cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam}, \name{Coq}
\cite{coqteam}, \name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems like
operators, logic variables, axioms, and theorems.

@@

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows. 
We briefly mention notation and syntax of the classical propositional
logic in section \ref{sec:preliminaries}.
In section \ref{sec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In section \ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
We listed some conclusions and future work in section
\ref{sec:conclusions}. At the end, we list some related works in 
proof reconstruction in section \ref{sec:related-work}.
The appendix \textsc{A} shows a complete example of the proof 
reconstruction workflow.

### 2017-09-11 paper/sections/metis-language-and-proofs.tex
[ paper ] implementation added, some proofs added.

\textit{Input.}~The \TPTP language --which includes the first-order
form (denoted by \name{fof}) and clause normal form (denoted by
\name{cnf}) formats~\cite{sutcliffe2009} -- is de facto input
standard language to encode problems for many \ATPs. The \TPTP
syntax describes a well-defined grammar to handle annotated formulas
with the following form,

@@

For instance, let us express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \vee \neg q)$
in \TPTP syntax as follows, declaring an \tptpaxiom $p$, and the
expected conclusion, the  \tptpconjecture
\verb!~ ((p & ~ p) | (q & ~ q))!.

@@

\textit{Output.}~\TSTP language is de facto output standard language
for derivations of ATPs~\cite{sutcliffe2004tstp}. A \TSTP derivation
is a directed acyclic graph, a proof tree, where each leaf is a
formula from the \TPTP input. A node is a formula inferred from the
parent formulas. The root is the final derived formula. Such a
derivation is a list of annotated formulas with the form,

@@

The \name{source} field is an inference record with the following
pattern,

@@

\begin{figure}
\centering
\begin{bprooftree}\tt
  \AxiomC{}
  \RightLabel{\footnotesize negate}
  \UnaryInfC{$\neg p$}
  \RightLabel{\footnotesize strip}
  \UnaryInfC{$\neg p$}
  \AxiomC{}
  \RightLabel{\footnotesize axiom}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize simplify}
  \BinaryInfC{$\bot$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$\bot$}
\end{bprooftree}
\caption{The \Metis' refutation tree for $p \vdash p$ from the
derivation in Fig.~\ref{fig:metis-proof-tstp}}
\label{fig:metis-example}
\end{figure}

>>>>

\textit{Input.}~The \TPTP language --which includes the first-order
form (denoted by \name{fof}) and clause normal form (denoted by
\name{cnf}) formats~\cite{sutcliffe2009} -- is de facto input
standard language to encode problems for many \ATPs. The \TPTP
syntax describes a well-defined grammar to handle annotated formulas
with the following form.

@@

For instance, let us express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \wedge \neg q)$
in \TPTP syntax as follows, declaring an \tptpaxiom $p$, and the
expected conclusion, the  \tptpconjecture
\verb!~ ((p & ~ p) | (q & ~ q))!.

@@

\textit{Output.}~\TSTP language is de facto output standard language
for derivations of ATPs~\cite{sutcliffe2004tstp}. A \TSTP derivation
is a directed acyclic graph, a proof tree, where each leaf is a
formula from the \TPTP input. A node is a formula inferred from the
parent formulas. The root is the final derived formula. Such a
derivation is a list of annotated formulas with the form.

@@

The \name{source} field is an inference record with the following
pattern.

@@

\begin{figure}
\centering
\begin{bprooftree}\tt
  \AxiomC{}
  \RightLabel{\footnotesize negate}
  \UnaryInfC{$\neg p$}
  \RightLabel{\footnotesize strip}
  \UnaryInfC{$\neg p$}
  \AxiomC{}
  \RightLabel{\footnotesize axiom}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$p$}
  \RightLabel{\footnotesize simplify}
  \BinaryInfC{$\bot$}
  \RightLabel{\footnotesize canonicalize}
  \UnaryInfC{$\bot$}
\end{bprooftree}
\caption{The \Metis' proof tree for $p \vdash p$ from the
derivation in Fig.~\ref{fig:metis-proof-tstp}}
\label{fig:metis-example}
\end{figure}
